---
title: "MESS-ing with joint Hill numbers"
author: "Andrew Rominger"
date: "4/22/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We're wondering if we can use joint Hill numbers to capture shared information found across multiple axes, such as abundances and $\pi$ values.  To explore this I use Chao's definition of a Hill number for trait differences and apply it to differences in (abundance, $\pi$) space.  Hill numbers are governed by a parameter $q$ to changes the contribution of rare versus common species (small $q$ mean rare species have more weight). Chao's functional trait Hill number is
$$
qH = \left(\sum_i^S\sum_j^S d_{ij} \left(\frac{p_ip_j}{Q}\right)^q\right)^{1/(1 - q)}
$$
where $d_{ij}$ is the matrix of distances between trait values of specie $i$ and species $j$; $p_k$ is the relative abundance of species $k$ (i.e. $n_k / J$) and Q is a normalization term defined as $\sum_i^S\sum_j^S d_{ij} p_ip_j$. This Hill number can be put in units of "effective species" like this:
$$
qD = \sqrt\frac{qF}{Q}.
$$

Now the trick is that instead of defining $d_{ij}$ in terms of traits, we can simply calculate distances between the abundances and $\pi$ values of different species, i.e. let $d_{ij} = \sqrt{(n_i - n_j)^2 + (\pi_i - \pi_j)^2}$.  Call Hill numbers from this type of distance "joint Hill numbers."

Below I code up the Hill functions

```{r}
library(mvtnorm)
library(pika)
library(parallel)

# function to return the Hill number for a given q
qH <- function(v, a, q, Q) {
    # browser()
    if(q == 1) {
        return(exp(- sum(v * a / Q * log(a / Q))))
    } else {
        return(sum(v * (a / Q)^q)^(1 / (1 - q)))
    }
}

# function to return the generalized Hill numbers (i.e. effective spp) for given q's
qD <- function(v, a, q) {
    Q <- sum(v * a)
    
    hh <- sapply(q, function(qi) {
        qH(v, a, qi, Q)
    })
    
    return(sqrt(hh / Q))
}
```

And now run an experiment calculating the joint Hill numbers for two arbitrary attributes.  I simulate scenarios where these attributes have a 0.75 covariance (look for `xcor` in the code below) and where they have 0 covariance (look for `xnoc`).

```{r, cache=TRUE}
# explore distribution of generalized Hill numbers for correlated and uncorrelated values
diffCors <- mclapply(1:200, mc.cores = 10, FUN = function(i) {
    S <- 1000
    xcor <- rmvnorm(S, rep(0, 2), matrix(c(1, 0.75, 0.75, 1), nrow = 2))
    xnoc <- rmvnorm(S, rep(0, 2), matrix(c(1, 0, 0, 1), nrow = 2))
    
    n <- rfish(S, 0.05)
    pij <- outer(n, n, '*') / (sum(n)^2)
    
    dijCor <- as.matrix(dist(xcor))
    dijNoc <- as.matrix(dist(xnoc))
    
    o <- c(qD(dijCor, pij, 1:4), qD(dijNoc, pij, 1:4))
    names(o) <- paste(rep(c('cor', 'noc'), each = 4), 1:4, sep = '_')
    
    return(o)
})

diffCors <- do.call(rbind, diffCors)
```


Now we look at generalized joint Hill numbers (across multiple $q$) for correlated and uncorrelated attributes.  You can see that, sadly, the Hill numbers completely overlap for both the correlated and uncorrelated cases.

```{r, fig.width=4, fig.height=4, fig.align='center'}
diffSum <- apply(diffCors, 2, quantile, probs = c(0.025, 0.5, 0.975))

par(mar = c(3, 3, 0, 0) + 0.5, mgp = c(2, 0.75, 0))

plot(diffSum[2, 1:4], ylim = range(diffSum), type = 'l', lwd = 3, 
     col = hsv(0, alpha = 0.5),
     panel.first = {
         polygon(c(1:4, 4:1), c(diffSum[1, 1:4], diffSum[3, 4:1]), 
                 col = hsv(0, alpha = 0.25))
         polygon(c(1:4, 4:1), c(diffSum[1, 5:8], diffSum[3, 8:5]), 
                 col = hsv(0.7, alpha = 0.25))
     }, 
     xlab = 'q', ylab = 'Generalized Hill')

points(diffSum[2, 5:8], ylim = range(diffSum), type = 'l', lwd = 3, 
       col = hsv(0.7, alpha = 0.5))

legend('topright', legend = c('correlated', 'uncorrelated'), lty = 1, lwd = 2, 
       col = hsv(c(0, 0.7), alpha = 0.5), bty = 'n')
```
